Improving Haskell Theory Exploration

### 2015-12-15 TransferReport/implementation.tex
Added Gantt chart and a little extra to evaluation

\iffalse TODO:
  A: Conceptually: - implementation follows the inductive/recursive structure of the previous definitions
e.g. definition of the rose tree is naturally represented as Haskell datatype definition

B Module-by-module road map
  - Treatment of Haskell package system

C - External software
\fi

@@

\subsection{\textsc{ASTPlugin}}
\label{sec:astplugin}

The GHC compiler provides mechanisms for parsing Haskell source code and converting it to Core. It also includes a \emph{renaming} transformation which makes all names unique (either by prefixing them with the name of the module which defines them, or by suffixing the name with a number. This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.

@@

  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc. Since Core does not make these distinctions, names may become ambiguous. For example, a type parameter \hs{t} may be confused with a function argument \hs{t}. To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder $\CLam\ \hs{"\$dOrd"}$ in Core. This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.

@@

The result of building a Haskell package with \textsc{AstPlugin} is a database of Haskell definitions, similar in some respects to \textsc{Hoogle}. Definitions are indexed by a combination of their package name, module name and binding name. The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name, which allows dependencies to be discovered. Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.

@@

As described in \S \ref{} We have a tool which annotates each definition with its dependencies, i.e. those (package, module, name) combinations which it references. Definitions are then topologically sorted into dependency order, as this is required for our recurrent clustering technique.

\subsection{Recurrent Clustering}

\iffalse TODO: maybe focus on the ``interesting cases'', and defer the nitty-gritty of extending the environment, etc. to the implementation section? \fi

Our implementation is a rather direct translation of the algorithm described in \S \ref{sec:contributions} into Haskell. At the top level, we send named expressions to \hs{recurrentCluster}, topologically sorted: if an element contains multiple \hs{(Global, Expr)} pairs, they are mutually-recursive:

\begin{haskell}
recurrentCluster :: [[(Global, Expr)]] -> [[Global]]
recurrentCluster = go ([],[])
  where go (fs, db) []       = db
        go (fs, db) (es:ess) = let fs' = fs ++ map (extract db) es
                                   db' = kMeans fs'
                                in go (fs', db') ess
        extract db (i, e) = (i, rt db e)
\end{haskell}

The overall result of the \hs{recurrentCluster} function is a list of clusters, containing the IDs of their elements. These are obtained by interleaving clustering (the \hs{kMeans} function) and feature extraction (the \hs{rt} function).

The feature extraction itself contains two parts: first, syntax trees matching the grammar in Figure \ref{fig:coresyntax} are converted into \hs{RoseTree}s with features (\hs{Float}s) on their \hs{Node}s:

\begin{haskell}
data RoseTree = Node Feature [RoseTree]

eRT :: [Local] -> [[Global]] -> Expr -> RoseTree
eRT env db e = case e of
  ...
\end{haskell}

The easiest branches to handle are literals and types, which we represent using particular feature values (\hs{sLITNUM}, \hs{sLITSTR} and \hs{sTYPE}):

\begin{haskell}
  Lit (LitNum _) -> Node sLITNUM []
  Lit (LitStr _) -> Node sLITSTR []
  Type           -> Node sTYPE   []
\end{haskell}

Function application simply recurses into both sub-expressions:

\begin{haskell}
  App e1 e2 -> Node sAPP [eRT env db e1,
                          eRT env db e2]
\end{haskell}

To look up variables locally and globally, we use the \hs{lookupL} and \hs{lookupG} functions, respectively. These return the index containing their argument, if found. In the case of \hs{lookupL}, this acts as a de Bruijn index to give alpha-equivalent terms equal feature vectors. For \hs{lookupG}, this is the ID of the cluster it appears in; this ensures that references to similar expressions result in similar features:

\begin{haskell}
  Var (Global i) -> Node (lookupG db  i) []
  Var (Local  i) -> Node (lookupL env i) []
\end{haskell}

Finally, when we traverse binders we must extend the environment \hs{env}:

\begin{haskell}
  Lam  i  e     -> Node sLAM [eRT (i:env) db e]
  Let  bs e     -> Node sLET (map (bRT (ids bs:env) db) bs ++
                                   eRT (ids bs:env) db  e
  Case e i alts -> Node sCASE (eRT    env  db  e :
                          map (aRT (i:env) db) alts)
\end{haskell}

Patterns and bindings are handled in a similar way:

\begin{haskell}
aRT :: [Local] -> [[Global]] -> Alt -> RoseTree
aRT env db alt = case alt of
  (DataAlt _, vs, e) -> eRT (vs ++ env) db e
  (LitAlt  _, _,  e) -> eRT env db e
  (Default,   _,  e) -> eRT env db e

bRT :: [Local] -> [[Global]] -> Bind -> RoseTree
bRT env db b = case b of
  NonRec i e -> eRT env db e
  Rec es     -> Node sREC (map (eRT env db . snd) es)
\end{haskell}

The result of \hs{eRT} is a \hs{RoseTree} whose branching structure mimics that of our original expression. We next need to convert this to a \emph{matrix} of features, which we do by converting each level of the \hs{RoseTree} into a row of the matrix (using \hs{pad} to ensure a consistent size). Finally we turn the matrix into a \emph{feature vector} by concatenating the rows:

\begin{haskell}
level :: Int -> RoseTree -> [Feature]
level 0 (Node x xs) = [x]
level n (Node x xs) = concatMap (level (n-1)) xs

rt :: [[Global]] -> Expr -> [Feature]
rt db e = concat (pad cols (map (`level` eRT [] db e) [0..rows]))
\end{haskell}

Notice that this algorithm contains several parameters, including \hs{rows} and \hs{cols} which determine how to truncate the matrices (defaults are 30). The \hs{kMeans} function also contains a parameter for the cluster number; we set this as $\sqrt{n}$ where $n$ is the number of feature vectors being clustered.

\iffalse TODO: How does this section differ from Algorithm in Contributions? \fi
\iffalse TODO: Make this a concrete presentation of the abstract ideas from the Algorithm section, e.g. this is how we get the dependencies (AST plugin, etc.); this is a case-by-case of how we perform feature extraction; this is how we perform k-means (Weka), etc. \fi

Most of the work carried out so far has been either preliminary investigations, to determine if a particular approach is worth pursuing; or at an infrastructure level to support experimentation.

\subsection{\textsc{ML4HS}}
\label{sec:ml4hs}

\textsc{ML4HS} is our top-level theory exploration framework, combining \textsc{AstPlugin} and \textsc{MLSpec}, described below, with an implementation of our recurrent clustering algorithm and the \textsc{Weka} machine learning library.

@@

When investigating heuristic methods like those of machine learning, it is important to use as much realistic data as possible to predict the system's performance on real tasks. One bottleneck for theory exploration is the lack of theories which are available to explore: we must manually select a set of terms to explore, and sometimes specify variables too.

Our \textsc{MLSpec} tool can automatically build theories suitable for exploration by \qspec{}, based on the databases constructed by \textsc{AstPlugin} described above. Features of \textsc{MLSpec} include:

\begin{itemize}
  \item{Monomorphising}: Given values of polymorphic type, e.g. \hs{safeHead :: forall t. [t] -> Maybe t} and \hs{[] :: forall t. [t]}, a testing-based system like \qspec{} is unable to evaluate expressions such as \hs{safeHead [] :: forall t. Maybe t} without instantiating the variable \hs{t} to a specific type. Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}. We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g. \hs{forall c. c Bool -> c Bool}) or incompatible class constraints (e.g. \hs{forall t. IsString t => t}).

  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity. There is still the possibility that multiple packages will declare modules of the same name, in which case the exploration process will abort.

@@

  \item{Sandboxing}: \textsc{MLSpec} is built on top of \texttt{nix-eval}, which allows the theory being explored to reference packages which aren't yet installed on the system.

@@

\subsection{\texttt{nix-eval}}
\label{sec:nixeval}

To facilitate theorem proving and machine learning, our theory exploration experiments require access to the \emph{definitions} of the terms under investigation. This is difficult, as many terms are functions, which Haskell does not let us inspect. This limitation can be worked around by programming at a meta-level: manipulating \emph{representations} of Haskell expressions, rather than the expressions themselves. To retain our ability to \emph{evaluate} expressions, we then need a mechanism to transform these representations into the expressions they represent; this usually takes the form of an \texttt{eval} function.

\texttt{eval} is a common feature in languages such as Python and Javascript, where extra source files can be imported dynamically. In that case \texttt{eval} can be implemented easily by representing expressions as strings, then treating those strings as if they were the content of a file being imported.

On the other hand, Haskell does not natively support dynamic importing of extra source files, making implementation of \texttt{eval} trickier. Comprehensive implementations do exist, such as \texttt{hint}, which are effectively wrappers around GHC, but all are limited to using the Haskell modules which are already installed on the system. Since we want to explore \emph{arbitrary} Haskell code, this is not enough for our purposes.

We have implemented a library called \texttt{nix-eval}, which provides an \hs{eval} function for Haskell expressions which may reference packages that are not installed on the system. These packages will be automatically downloaded and installed into a sandbox when a call to \hs{eval} is forced.

\texttt{nix-eval} is build on top of the Nix package manager and GHC. Nix provides a language for defining packages, a \texttt{cabal2nix} tool for generating Nix package definitions from those of Haskell's Cabal tool, and a \texttt{nix-shell} tool for evaluating arbitrary shell commands in a sandbox containing arbitrary packages.

The internal representation of \texttt{nix-eval} includes a list of package names, a list of module names, a list of GHC options and a string containing Haskell code. When evaluated, \texttt{nix-shell} is invoked to create a sandbox containing an instance of GHC and the Haskell packages listed in the expression. GHC is invoked in this sandbox using the \texttt{runhaskell} command, along with any options given in the expression. The code to be evaluated is prepended with \hs{import} statements for each required module, and wrapped in a simple \hs{main} function for printing the result of its evaluation to standard output. The expression, therefore, must have type \hs{String}; although this is not enforced statically.

The type of the \hs{eval} function is \hs{Expr -> IO (Maybe String)}, where \hs{Expr} is the type of expressions described above. The \hs{IO} wrapper comes from our invocation of external tools, and \hs{Maybe String} allows us to represent failure in a clean way. The restriction of expressions to \hs{String}s is not onerous; some concrete type is required, to communicate results from the GHC invocation, but we consider the particular choice of encoding to be out of scope for \texttt{nix-eval}. It would be a straightforward exercise to wrap our \hs{eval} function in a generic encoding mechanism, e.g. using JSON.

\subsection{Circular Convolution}
\label{sec:circularconvolution}

We have investigated the reduction to a fixed-size of tree structures of the following form (where \hs{FV} denotes a feature vector):

\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
data Tree = Leaf FV
          | Branch FV Tree Tree
\end{lstlisting}

Just like in the non-recursive case, the simplest way to reduce our dimensionality is to truncate. To reduce an arbitrary \hs{t :: Tree} to a maximum depth \hs{d > 0}, we can use \hs{reduceD d t} where:

\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
reduceD 1 (Branch fv l r) = Leaf   fv
reduceD n (Leaf   fv)     = Leaf   fv
reduceD n (Branch fv l r) = Branch fv (reduceD (n-1) l)
                                      (reduceD (n-1) r)
\end{lstlisting}

This is simple, but suffers two problems:

\begin{itemize}
 \item We must choose a conservatively large \hs{d}, since we're throwing away information
 \item The number of feature vectors grows exponentially as \hs{d} increases
\end{itemize}

The first problem can't be avoided when truncating, but the second can be mitigated by truncating the \emph{width} of the tree as well, to some \hs{w > 0}. One way to truncate the width is tabulating the tree's feature vectors, then truncating the table to \hs{w} $\times$ \hs{d}, as ML4PG does.

\iffalse

We can do this with \hs{tabulate w d t}:

\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
tabulate w d t = take d (map (take w) (tabulate' t))

tabulate' (Leaf   fv)     = [fv] : []
tabulate' (Branch fv l r) = [fv] : merge (tabulate' l) (tabulate' r)

merge    []     ys  = ys
merge    xs     []  = xs
merge (x:xs) (y:ys) = (x ++ y) : merge xs ys
\end{lstlisting}

\fi

Rather than truncating our trees, we can \emph{combine} the feature vectors of leaves into their parents, and hence \emph{fold} the tree up to any finite depth. Following \citep{zanzotto2012distributed} we have investigated the use of \emph{circular convolution} (\hs{cc}) as our combining function. Hence, to reduce \hs{t :: Tree} to a single feature vector, we can use \hs{reduceC t} where:

\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
reduceC (Leaf fv)       = fv
reduceC (Branch fv l r) = cc fv (cc (reduceC l) (reduceC r))
\end{lstlisting}

Circular convolution has the following desirable properties:

\begin{description}

  \item{Non-commutative}: \hs{cc a b} $\not\approx$ \hs{cc b a}, hence we can distinguish between \hs{Branch fv a b} and \hs{Branch fv b a}

  \item{Non-associative}: \hs{cc a (cc b c)} $\not\approx$ \hs{cc (cc a b) c}, hence we can distinguish between \hs{Branch v1 a (Branch v2 b c)} and \hs{Branch v1 (Branch v2 a b) c)}.

\end{description}

Here we use $\not\approx$ to represent that these quantities differ \emph{with high probability}, based on empirical testing. This is the best we can do, since the type of combining functions \hs{(FV, FV) -> FV} has a co-domain strictly smaller than its domain, and hence it must discard some information. In the case of circular convolution, the remaining information is spread out among the bits in the resulting feature, which is known as a \emph{distributed representation} \citep{conf/ijcai/Plate91}.

Distributed representations mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.

\subsubsection{Extracting Features from XML}

This method of folding with circular convolution has been implemented in the Haskell program \textsc{Tree Features} \footnote{Available online at \href{http://chriswarbo.net/git/tree-features}{http://chriswarbo.net/tree-features}. The application parses arbitrary trees of XML and uses binary features, ie. feature vectors are bit vectors.

To calculate the initial feature of an XML element, before any folding occurs, we concatenate its name and attributes to produce a string \hss}. The feature is then:

\begin{equation}
feature(n) = 2^{\md5(s) \pmod{L}}
\end{equation}

Where $\md5$ is the MD5 hash algorithm and $L$ is the length of the desired vector, given as a commandline argument. The result of $feature(s)$ is a bit vector of length $L$, containing only a single $1$. Due to the use of a hashing function, the position of that $1$ can be deterministically calculated from the input $s$, yet the \emph{a priori} probability of each position is effectively uniform.

In other words, the hash introduces unwanted patterns which are \emph{theoretically} learnable, but in practice a strong hash can be treated as irreversible and hence unlearnable.

\subsubsection{Application to Coq}

Prior to 2014-09-08, the Coq proof assistant provided a rudimentary XML import/export feature, which we can use to access tree structures for learning. We achieve this by using the \coq{external} primitive of the Ltac tactic language: \coq{external "treefeatures" "32" t1 t2 t3.} will generate an XML tree representing the Coq terms \coq{t1}, \coq{t2} and \coq{t3}, and send it to the standard input of a \texttt{treefeatures} invocation, using a vector length argument of 32.

Coq will also interpret the standard output of the command, to generate terms and tactics. This functionality isn't yet used by \textsc{Tree Features}, but it is clear that a feedback loop can be constructed, allowing the construction of powerful \textsc{Ltac} tactics which invoke external machine learning systems.

>>>>

@@

\subsection{\textsc{AST Plugin}}
\label{sec:astplugin}

The GHC compiler provides mechanisms for parsing Haskell source code and converting it to Core. It also includes a \emph{renaming} transformation, which resolves global identifiers into a canonical form. This allows us to spot repeated use of a term, across multiple modules and packages, with a simple syntactic equality check.

@@

  \item{Unified namespace}: Haskell has distinct namespaces for values, types, data constructors, etc. Since Core does not make these distinctions, names may become ambiguous. For example, a type parameter \hs{t} may be confused with a function argument \hs{t}. To prevent this, overlapping namespaces are distinguished by prefices which are distinct from the available names; for example a type class constraint \hs{Ord t} may give rise to a binder $\CLam\ \hs{"\$dOrd"}$ in Core, which is guaranteed not to conflict since this name would be invalid in Haskell. This causes difficulties when looking up names, as these prefixed forms do not easily map back to the Haskell source.

@@

The result of building a Haskell package with \textsc{AstPlugin} enabled is a database of Haskell definitions, similar in some respects to \textsc{Hoogle} \citep{mitchell2008hoogle}. Definitions are indexed by a combination of their package name, module name and binding name. The definitions themselves are s-expressions representing the Core AST, with non-local references replaced by a combination of package name, module name and binding name, which makes it trivial to look up references in the database. Each definition also has an associated arity and type, obtained during the post-processing step mentioned above.

@@

As described in \S \ref{sec:symbolstofeatures}, we must topologically sort the output of \textsc{AstPlugin} in order for our recurrent clustering to be well-founded. Since our database keys (containing the package, module and binding names, as described above) match our representation of non-local references, it is simple to walk each syntax tree to obtain the set of references it makes. In addition, the resulting set of (identifier, list-of-referenced-identifiers) pairs exactly matches the (vertex, list-of-successor-vertices) format used to represent directed graphs by the popular \hs{containers} library which ships with GHC. This provides an implementation of topological sort for strongly connected components, which we use as-is. A simple shell script loops through these SCCs, invoking the recurrent clustering component for each and appending the resulting features and clusters to the database.

\subsection{Feature Extraction}

The implementation of our feature extraction algorithm is a rather direct translation of the description given in \S \ref{sec:contributions} into Haskell. We parse the s-expressions generated by \textsc{AstPlugin} into algebraic data types which correspond directly to the definitions in Figure \ref{fig:coresyntax}; this is routine, so we omit the details for brevity. Similarly, we can represent rose trees with a datatype corresponding to the definition given in \S \ref{sec:expressionstovectors}:

\begin{haskell}
data RoseTree = Node Feature [RoseTree]
\end{haskell}

For simplicity we use the representation \hs{Feature = Int}, as we do not have fractional values. Since we represent the symbols $expr$, $id$, etc. from Figure \ref{fig:coresyntax} using different datatypes, we cannot write one big definition of $toTree$ or $\phi$ which works on all tokens. Each case shown in Figure \ref{fig:totree} and equations \ref{eq:feature}, \ref{eq:localfeature} and \ref{eq:globalfeature} appears in the implementation, although they are spread across several functions.

To support looking up local identifiers, our implementation of $toTree$ takes a context as argument, extending it as required. As an example of the complexity this adds, here is the \CCase\ branch of $toTree$:

\begin{lstlisting}[language=Haskell, xleftmargin=0pt, xrightmargin=0pt]
toTree :: Context -> Expr -> RoseTree
toTree ctx x = case x of
  ...
  Case e l as -> Node fCase (toTree ctx e : map (toTreeAlt (l:ctx)) as)
  ...
\end{lstlisting}

Breaking this down we can see \hs{fCase} representing the value of $\feature{\CCase}$, and a list of sub-trees defined in parentheses. The first subtree is a straightforward recursive call in an unmodified context: \mbox{\hs{toTree ctx e}}. The rest of the list is formed by applying the function \mbox{\hs{toTreeAlt (l:ctx)}} to each element of the list \hs{as} of \CAlt\ clauses.

The \hs{toTreeAlt} function contains those cases of $toTree$ which handle symbols in $alt$. We prepend the identifier \hs{l} to the context, to get the extended context \hs{l:ctx}. This is because \hs{l} will be bound the value of \hs{e}, in order to avoid re-computing its value several times.

The other clauses are handled in a similar way. The trickiest is the \CLet\ clause, since the local identifiers aren't directly available; we must extract them from their \CRec, \CNonRec\ and \CBind\ wrappers first, which we do using helper functions.

As shown above, the values from equations \ref{eq:feature} are encoded directly in $toTree$. For $phi(l \in \mathcal{L})$ we use standard Haskell functions to look up the required indices in the context:

\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
phiL :: Context -> Local -> Feature
phiL ctx x = case elemIndex x ctx of
  Nothing -> error (concat ["Local '", show x,
                            "' not in context '",
                            show ctx, "'"])
  Just i  -> (2 * alpha) + i
\end{lstlisting}

As explained in \S \ref{sec:symbolstofeatures}, local identifiers should always exist in the context. If this precondition doesn't hold, we abort the program with an error rather than continuing.

Global identifiers are kept as-is until we have access to the clusters from the last iteration. This takes place outside Haskell, using the \hs{jq} data processing tool.

Our implementation of $level$ exactly matches equation \ref{eq:level}:

\begin{haskell}
level :: Int -> RoseTree -> [[Feature]]
level 1 (Node f _)  = [f]
level n (Node _ ts) = concatMap (level (n-1)) ts
\end{haskell}

To produce feature vectors, we do not directly construct the matrix; instead we generate the rows and concatenate them together in one step, using the \hs{concatMap} function:

\begin{lstlisting}[language=Haskell, xleftmargin=0.1\textwidth, xrightmargin=0.1\textwidth]
featureVec :: Expr -> [Feature]
featureVec e = concatMap (\m -> pad (level m tree)) [1..r]
  where tree   = toTree [] e
        pad xs = take c (xs ++ repeat 0)
\end{lstlisting}

By providing \hs{featureVec} with the latest set of clusters, read from the \textsc{AstPlugin} database, we turn Core expressions into feature vectors, which are appended to the database.

We use the Weka system to perform our k-means clustering, as it is widely used, including by ML4PG. We select all feature vectors from our database, and write them in CSV format for Weka to process. The Weka CLI command is invoked, which appends a cluster number to each of these feature vectors; we read these off and append them to the database. As long as more SCCs remain unprocessed, we keep looping this process, using the database to communicate between the feature extraction and clustering phases.

@@

We cannot supply these clusters as-is to \qspec{}, since it must be provided with a \emph{signature}. These are constructed by our \textsc{MLSpec} tool, using information from the \textsc{AstPlugin} database. Tasks performed by \textsc{MLSpec} include:

\begin{itemize}
  \item{Monomorphising}: Given values of polymorphic type, e.g. \hs{safeHead :: forall t. [t] -> Maybe t} and \hs{[] :: forall t. [t]}, a testing-based system like \qspec{} is unable to evaluate these expressions without instantiating the variable \hs{t} to a specific type. Such an instantiation is called \emph{monomorphising}, and in the case of \textsc{MLSpec} we build on previous work in \qcheck{} by attempting to instantiate all type variables to \hs{Integer}. We discard those cases where this is invalid, such as variable \emph{type constructors} (e.g. \hs{forall c. c Bool -> c Bool}) or incompatible class constraints (e.g. \hs{forall t. IsString t => t}).

  \item{Qualification}: All names are \emph{qualified} (prefixed by their module's name), to avoid most ambiguity. There is still the possibility that multiple packages will declare modules of the same name, although this is rare as it causes problems for any Haskell programmer trying to use those modules. In such cases the exploration process simply aborts.

@@

  \item{Sandboxing}: One difficulty with Haskell's packaging infrastructure is that all required packages and modules must be provided up-front, usually by specification in a Cabal file. Since \textsc{MLSpec} builds signatures \emph{dynamically}, depending on the cluster information it is given, we do not know what packages it may need. To work around this problem, \textsc{MLSpec} invokes \qspec{} for each cluster using a library we have built called \texttt{nix-eval}. This provides an \texttt{eval} function, like those commonly found in dynamic languages such as Python and Javascript, for evaluating Haskell expressions. The key feature of \texttt{nix-eval} is that these Haskell expressions may reference packages that are not installed on the system. When such expressions are evaluated, these packages will be automatically downloaded and installed into a sandbox using the Nix package manager, and GHC will be invoked in this sandbox to perform the evaluation.

@@

### 2015-12-15 TransferReport/intro.tex
Recurrent clustering section
Added Gantt chart and a little extra to evaluation

As computers and software become more sophisticated, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity. The \emph{functional programming} paradigm has been proposed for addressing these issues \citep{hughes1989functional}, by constructing programs which are more amenable to mathematical analysis. \iffalse TODO: Types? \fi For example, in pure functional programming all values are \emph{immutable}: defined once and never changed. Hence there is no way for a value to be altered between the point it is introduced and the point it is used, unlike in many \emph{imperative} languages where we may have to search the whole program to ensure the value is not altered by any intermediate code. Similarly, pure functions cannot depend on any state other than their arguments, and hence will always produce repeatable results.

@@

\begin{enumerate}
  \item The application of machine learning algorithms to theory exploration, for intelligently discovering interesting sub-sets of large signatures, which are more tractable to explore.
  \item A novel feature extraction method for transforming Haskell expressions into a form amenable to off-the-shelf learning algorithms.
  \item An implementation of these feature extraction and theory exploration approaches.
  \item A comparison of our methods with existing approaches, both for theory exploration in Haskell, and for machine learning in other languages.
\end{enumerate}

We begin in \S \ref{sec:background} by providing a formal context for analysing Haskell expressions (\S \ref{sec:haskell}) and describe the \qspec{} theory exploration system (\S \ref{sec:theoryexploration}). We give a brief overview of testing approaches and how they relate to Haskell (\S \ref{sec:quickcheck}), as well as the machine learning approaches we are building on (\S \ref{sec:featureextraction}). We discuss our contributions in more depth in \S \ref{sec:contributions}, and provide implementation details \S \ref{sec:implementation}. A variety of related work is surveyed in \S \ref{sec:related}, and we give several potential directions for future research in \S \ref{sec:future}.

>>>>

As computers and software become more sophisticated, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity. The \emph{functional programming} paradigm has been proposed for addressing these issues \citep{hughes1989functional}, by constructing programs which are more amenable to mathematical analysis. For example, in pure functional programming all values are \emph{immutable}: defined once and never changed. Hence there is no way for a value to be altered between the point it is introduced and the point it is used, unlike in many \emph{imperative} languages where we may have to search the whole program to ensure the value is not altered by any intermediate code. Similarly, the results of pure functions cannot depend on any state other than their arguments, and hence will always produce repeatable results. By making state implicit in this way, powerful type systems can be used to constrain the behaviour of programs, and to give a rich, composable structure to data.

@@

\begin{enumerate}
  \item The application of machine learning algorithms to theory exploration, for intelligently discovering interesting sub-sets of Haskell libraries, which are more tractable to explore.
  \item A novel feature extraction method for transforming Haskell expressions into a form amenable to off-the-shelf learning algorithms.
  \item An implementation of these feature extraction and theory exploration approaches.
  \item A comparison of our methods with existing approaches, both for theory exploration in Haskell, and for machine learning in other languages.
\end{enumerate}

We begin in \S \ref{sec:background} by providing a formal context for analysing Haskell expressions (\S \ref{sec:haskell}) and describe the \qspec{} theory exploration system (\S \ref{sec:theoryexploration}). We give a brief overview of testing approaches and how they relate to Haskell (\S \ref{sec:quickcheck}), as well as the machine learning approaches we are building on (\S \ref{sec:featureextraction}). We discuss our contributions in more depth in \S \ref{sec:contributions}, and provide implementation details in \S \ref{sec:implementation}. A variety of related work is surveyed in \S \ref{sec:related}, we briefly evaluate our implementations in \S \ref{sec:evaluation} and give several potential directions for future research in \S \ref{sec:future}.

### 2015-12-15 TransferReport/related.tex
Added Gantt chart and a little extra to evaluation

@@

Recently, a variety of alternative algorithms have also been investigated, for example the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \citep{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses. Visibility is essentially a dependency graph of which theorems were used in the proofs of which other theorems (although the theorems are actually represented as abstract sets of features). To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a naive Bayes algorithm.

@@

>>>>

\subsection{Theory Exploration}

We briefly described theory exploration in \S \ref{sec:theoryexploration}, as the task of discovering \emph{new} theorems in a software or proof library, rather than proving/disproving user-provided statements. The idea was first introduced in the \textsc{Theorema} \cite{buchberger2000theory} system of Buchberger. This provided an interactive environment, similar to computer algebra systems and interactive theorem provers. In this setting, many of our concerns such as the generation of values and deciding which properties to explore are simply delegated to the user; the software would check for correctness, store results and perform searches; again, similar to interactive theorem provers.

Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery. By removing user interaction, these concerns about directing search must be solved by algorithms. As well as \qspec{} and \hspec{} in Haskell, automated theory exploration has been applied to Isabelle \citep{Montano-Rivas.McCasland.Dixon.ea:2012, johansson2009isacosy, Hipster}.

We have focused our attention on \qspec{}, although it does not actually \emph{prove} its results, and hence may not be considered a theory exploration system on its own. However, it does form a vital component of \hspec{}, which uses off-the-shelf automated theorem provers (ATPs) to verify \qspec{}'s conjectures, forming a complete theory exploration system as well as a capable inductive theorem prover (by exploiting theory exploration for lemma generation) \cite{claessen2013automating}. Due to \hspec{}'s use in \textsc{Hipster}, improvements to \qspec{} also benefit work being pursued in Isabelle.

@@

Recently, a variety of alternative algorithms have also been investigated, for example the \textsc{MaSH} algorithm (Machine Learning for SledgeHammer) \citep{kuhlwein2013mash} uses the ``visibility'' of one theorem from another to determine the relevance of clauses. Visibility is essentially a dependency graph of which theorems were used in the proofs of which other theorems (although the theorems are actually represented as abstract sets of features). To select relevant clauses for a goal, the set of clauses which are visible from the goal's components is generated; this is further reduced by (an efficient approximation of) a na\"{\i}ve Bayes algorithm.

@@

\iffalse
TODO: Theory Exploration?
\citep{johansson2009isacosy}
\fi
